# 4. Clustering and classification

*This file describes the work and results of the fourth week a. k. a. "Clustering and classification" of the IDOS2023 course.*

### 4.1 Information on data set

The **Boston** data set from the **MASS** package in R consists of information on different characteristics for suburbs in Boston, Massachusetts, US. Variables include, amongst others:

- *crim*: per capita crime rate per town, 
- *zn*: proportion of residential land zoned for lots over 25,000 sq.ft. (2322.576 sq. meter),
- *nox*: nitrogen oxides concentration (parts per 10 million),
- *chas*: Charles River dummy variable (= 1 if tract bounds river; 0 otherwise),
- *rad*: index of accessibility to radial highways. 

The data set contains 506 rows (observations) and 14 columns (variables), of which all are numerical values and none are characters. *chas* is a binary integer, and *rad* an integer number. More information on the data set, its variables the abbreviations can be found [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).

```{r, message = FALSE}
# access all packages needed in this chunk
library(MASS)
library(dplyr)
library(tidyverse)

# load the data
data("Boston")

# explore the dataset
str(Boston)
dim(Boston)
```


#### Graphical and numerical overview

The black-and-white plot matrix below shows that numerous variables are grouped in two groups, *e. g.*, *chas* (binary), *rad*, and *tax*. *crim* and *zn* indicate that many observations are 0 (*i. e.*, no crimes and no large residental home, respectively). Most of the variables are not normally distributed: the proportion of owner-occupied units built prior to 1940 (*age*) is high, as well as the amount of black people living in the suburbs (*black*). 

```{r, message = FALSE}
# access all packages needed in this chunk
library(dplyr)
library(tidyverse)

# show summaries of variables
summary(Boston)
# plot matrix of the variables
pairs(Boston)
# histograms
Boston %>% 
     gather() %>% 
     ggplot(aes(x=value)) + geom_histogram(binwidth = 1) + facet_wrap('key', scales='free')

```

When we look at the coloured correlation matrix below, we can see the correlations between the variables more clearly. Big circles show a strong correlation, whereas small show a weak or no correlation, also noticeable with faint colour. The blue and red circles indicate a positive and negative correlation, respectively. A very strong positive correlation can, *e. g.*, be seen between *rad* and *tax*. A strong negative collrelation can, *e. g.*, be seen between *age* and *dis* (weighted mean of distances to five Boston employment centers). 

```{r, message = FALSE}
# access all packages needed in this chunk
library(corrplot)
library(dplyr)

# calculate the correlation matrix and round it
cor_matrix <- cor(Boston) %>%
  round(digits = 2)
# visualise the correlation matrix
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)

```

### 4.2 Scale the data set and divide it

To perform a linear discriminant analysis, it is necessary to scale the data. For this, we subtract the column means from the corresponding columns and divide the difference with the standard deviation:

$$scaled(x) = \frac{x - mean(x)}{ sd(x)}$$

```{r}
# center and standardize variables
boston_scaled <- Boston %>%
  scale()
# summaries of the scaled variables
summary(boston_scaled)
sd(boston_scaled$zn)
sd(boston_scaled$age)
# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)
# change crim to numeric
boston_scaled$crim <- as.numeric(boston_scaled$crim)
# create a quantile vector of crim 
bins <- quantile(boston_scaled$crim)
# create a categorical variable 'crime'
labels <- c("low", "med_low", "med_high", "high")
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = labels)
# look at the table of the new factor crime
table(crime)
# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)
# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# divide data set in test and train
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)
# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)
# create train set
train <- boston_scaled[ind,]
# create test set 
test <- boston_scaled[-ind,]

```

When we look at the scaled data, we can see in the summary that the mean of all variables equals 0. Similarly, the standard deviation is 1 for all variables (not seen for all). 

In order to then predict what might happen in Boston's suburbs in the future, we need to know how well the model we will use works. For this, we split the original data set into a train (80% of the data) and test set (20 % of the data). We can then train the model with the train set and predict with the test set. 


### 4.3 Linear discriminant analysis

[Linear discriminant analysis](https://en.wikipedia.org/wiki/Linear_discriminant_analysis) is a statistical method that tries to find linear combinations of explanatory variables and group them in differences that are as large as possible. It weighs the explanatory variables (predictors), creates functions out of it (so-called linear discriminant functions, *i. e.*, LD1, LD2, LD3, see below) and distinguishes them as much as possible. 

```{r}
# crime = target variable, . = all other (explanatory) variables
lda.fit <- lda(crime ~ ., data = train)
# print the lda.fit object
lda.fit
# load function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  graphics::arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}
# target classes as numeric
classes <- as.numeric(train$crime)
# plot the lda results (select both lines and execute them at the same time!)
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

# save the correct classes from test data
correct_classes <- test$crime
# remove the crime variable from test data
test <- dplyr::select(test, -crime)

```

From the summary above ("Prior probabilities of groups"), we can see that based on the training data, 25 % of the data set belongs to the low group, 25 % to med_low, 24% to med_high and 26% to high, respectively. The proportion of trace shows the between-class variance in the different linear discriminant functions. Hence, 96.5% of the between-class variance is explained by the first linear discriminant function (LD1). The coefficients (of linear discriminants) indicate that *rad* (index of accessibility to radial highways) is very well represented in LD1 (4.06) compared to all other variables ranging around 0. This is confirmed by looking at the plot, where *rad* seems to be the only variable strongly influencing LD1. We can also see the groups of observations vary a lot in LD1 (x-axis), especially the *high* group clustered on the other end. LD2 (y-axis) does not show a discriminative power, so does not capture / group the differences in the explanatory variables well. 

### 4.4 Predictions

After training the model, we can now predict classes with the LDA model on the test data. If we look at the categorical accuracy, we can see that the accuracy of the predictions for *high* is highest, followed by *med_low*, *low*, and *med_high* (95%, 64%, 60%, and 42%, respectively). This is also evident in the cross-tabulation. Most *high* values were correctly predicted, only 1 was wrongly predicted as *med_high*.


```{r}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)
# cross tabulate the results
conf <- table(correct = correct_classes, predicted = lda.pred$class)
conf
# calculate precision
diag(conf) / rowSums(conf)

```


### 4.5 Distance measures and k-means clustering

To state whether objects are similar to one another or not, we can also measure distances. The most common distance measure is the [Euclidean distance](https://en.wikipedia.org/wiki/Euclidean_distance), which is the length of a straight line (distance) between two points and its x and y coordinate. 

[K-means clustering](https://en.wikipedia.org/wiki/K-means_clustering) is a commonly used clustering method to assign observations to groups (a. k. a. clusters) based on how similar they are. 

```{r, fig.height= 14, fig.width= 14, message = FALSe}
# reload Boston data set
library(MASS)
data("Boston")
# standardise data set
boston_scaled <- as.data.frame(scale(Boston))
boston_scaled$crim <- as.numeric(boston_scaled$crim)

# euclidean distance matrix
dist_eu <- dist(boston_scaled)
# look at the summary of the distances
summary(dist_eu)

# k-means clustering with 2 clusters
km <- kmeans(boston_scaled, centers = 2)
pairs(boston_scaled, col = km$cluster)
# k-means clustering with 3 clusters
km <- kmeans(boston_scaled, centers = 3)
pairs(boston_scaled, col = km$cluster)
# k-means clustering with 4 clusters
km <- kmeans(boston_scaled, centers = 4)
pairs(boston_scaled, col = km$cluster)
# k-means clustering with 5 clusters
km <- kmeans(boston_scaled, centers = 5)
pairs(boston_scaled, col = km$cluster)

```

From the different plots above, 4 clusters seem to represent the differences in the explanatory variables relatively well. Rather than manually trying to find the optimal k for the k-means, however, we can determine the number of clusters by looking at the changes of the total of within cluster sum of squares (WCSS). The optimal number of clusters is when we can see a radical drop in TWCSS (see below).

```{r}
# Work with the exercise in this chunk, step-by-step. Fix the R code!
# MASS, ggplot2 and Boston dataset are available
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <- kmeans(boston_scaled, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_scaled, col = km$cluster)


```




















