# Logistic regression

*This file describes the work and results of the third week a. k. a. "Logistic regression" of the IDOS2023 course.*

## Information on dataset

The original data set includes data from student performances in the subjects "Mathematics" and "Portuguese language" in secondary education (high school) in two Portuguese schools. It was collected through school reports and questionnaires, and gives information on the demographic and social background of the students, as well as student grades and data on the school they visit (click [here](https://archive.ics.uci.edu/dataset/320/student+performance) to see more information on the data set and access the data). One of the collected variables concerns the alcohol consumption of the students. 

The data set we use has 370 observations / rows with 35 variables / columns (see below). It combines the data from two student alcohol consumption data sets, where 'alc_use' gives the average of the weekly alcohol consumption from 1 - very low to 5- very high, and 'high_use' = TRUE when students consume more than a little amount of alcohol (> 2). 

```{r, message = FALSE}
# message = FALSE will not show any message from R

# Read and summarise file 
# load saved file from data wrangling exercise 
alc <- read.table(file='/home/ntriches/github_iods2023/IODS23/data/alc.csv', header=TRUE, sep = ",")
colnames(alc) # names of variables 
str(alc)      # structure of data
dim(alc)      # dimension of data

```

## Analysis 

## Aim

The aim of my analysis is to study the relationships between high / low alcohol consumption and the gender of the students ("sex", female / male), if they go out with friends often ("goout", 1 - very low to 5 - very high), if they miss school often ("absences", number of school absences) and how well they performed in their final grades ("G3", 0 to 20). My personal hypotheses are as follows:

1. Students identifying as female drink less (relatively) than students identifying as males. 
2. If the students go out with friends often, they have a higher alcohol consumption.
3. The higher the number of school absences, they higher the alcohol consumption.
4. The more the students drink, the lower their final grade is. 

## Numerical and graphical exploration 


The following figure...

Numerically and graphically explore the distributions of your chosen variables and their relationships with alcohol consumption (use for example cross-tabulations, bar plots and box plots). Comment on your findings and compare the results of your exploration to your previously stated hypotheses. (0-5 points)

```{r}
# load needed libraries 
library(dplyr); library(ggplot2)

# summary statistics by group
# differences between female / male and high / low alcohol consumption, mean of going out, mean of absences, mean of final grade
alc %>% 
  group_by(sex, high_use) %>% 
  summarise(count = n(), 
            mean_goout = mean(goout),
            mean_absences = mean(absences), 
            mean_grade = mean(G3)
            )

# boxplots
plot_alc_use_goout_by_sex <- alc %>%
  ggplot(aes(x = high_use, y = goout, col = sex)) +
  geom_boxplot()
plot_alc_use_goout_by_sex

plot_alc_use_absences_by_sex <- alc %>%
  ggplot(aes(x = high_use, y = absences, col = sex)) +
  geom_boxplot()
plot_alc_use_absences_by_sex

plot_alc_use_grades_by_sex <- alc %>%
  ggplot(aes(x = high_use, y = absences, col = sex)) +
  geom_boxplot()
plot_alc_use_grades_by_sex

```

1. 40% of males show a high alcohol consumption (70 TRUE / 175 total * 100). Compared to 21 % of the females showing a high alcohol consumption (41 TRUE / 195 total * 100), I can assume that my first hypothesis, that males drink more than females, is correct.
2. Females who have a high alcohol consumption go out more often (3.4) compared to those who go out less frequently (3). For males, the differences is even greater: Those going out often (3.9) drink more compared to those who go out less (2.7). I can therefore assume that my seconds hypothesis, that those students who go out often drink more, is correct.
3. Females seem to miss school more often than males but for both genders, the same trend is evident: They show a high use of alcohol with a higher number of absences. I can assume that my third hypothesis can be accepted.
4. Males who show a high alcohol consumption achieve lower grades in their final exam compared to those with low alcohol consumption (10.3 and 12.3, respectively). For females, the differences in the final grades are smaller and show an opposite trend: with high alcohol consumption, the final grades are slightly higher than with low alcohol consumption (11.8 and 11.4, respectively). My fourth hypothesis, that high alcohol consumption results in lower final grades, can therefore only be accepted for males. 

